%=========================================================================
% Start of first day of lines and parameterization of a line
%=========================================================================
\preClass{Lines}

\begin{problem}
\item A line with slope 3 goes through the point $(-2,5)$. Express the
  formula for the line in different ways.
  \begin{subproblem}
  \item Express the formula for the line in slope-intercept form.
    \vfill
  \item Express the formula for the line in point-slope form.
    \vfill
  \item Express the formula for the line in standard form.
    \vfill
  \end{subproblem}

  \clearpage

\item A turtle starts at the origin. It moves at a constant speed of 2
  meters per minute. It is facing in a direction $\frac{\pi}{4}$
  radians from the $x$-axis and is moving in the first
  quadrant. Determine its position at any time.

  \clearpage

\end{problem}


\actTitle{Parameterization of a Line}
\begin{problem}
\item Find the equation for a line that goes through the point
  $<1,5,-2>$  and is in the direction  $<4,-2,6>$.

  \vfill

\item Suppose that an object's position is given by the formula you
  had in the previous question. Determine its velocity.
  \sideNote{Determining velocity from position is no different in the
    context of vectors.}

  \vfill

\item Suppose that an object has a constant velocity given my
  $<4,-2,6>$ m/sec. If the object starts at the point $<1,5,-2>$
  determine its position at any time.

  \vfill

\end{problem}


\postClass

\begin{problem}
\item Briefly state two ideas from today's class.
  \begin{itemize}
  \item 
  \item 
  \end{itemize}
\item 
  \begin{subproblem}
    \item
  \end{subproblem}
\end{problem}




%%% Local Variables:
%%% mode: latex
%%% TeX-master: "labManual"
%%% End:
